\tit

### 2021-02-17 paper.tex
proving and transporting properties

The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\api}.

@@

\figrefdef{predicates}{Predicate operations}{
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{-2ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{17em}
%% Transform codomain: %\small
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{20em}
%% Transform domain: %\small
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{-2ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which also includes language complement (\AF ∁).
\rnc\source{Calculus}
The language representations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}) resulting in broader coverage with less code.
The relevant \AF{ν} and \AF{δ} lemmas in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.

@@

\subsectionl{Predicate Algebra}

@@

There is still more algebraic structure to be found and exploited.
When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Languages, Predicates, and Types}.
In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Languages, Predicates, and Types}.
The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
The proofs of these properties, while not terribly complex, are beyond the scope of this paper but are available in the paper's Agda source code.\notefoot{There will be short and long forms of this paper. Perhaps the long form will contain narrated parts of the monoid semiring proof.}

Not only do types form a closed semiring, the \emph{only} properties needed from types in the monoid semiring come from the laws of closed semirings.
We can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
\out{We will explore this generality in \secref{Beyond Predicates}.}
As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.

\subsectionl{Transporting algebraic properties}

>>>>

The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.

@@

\figrefdef{predicates}{Predicate operations}{
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{-2ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{17em}
%% Transform codomain: %\small
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{20em}
%% Transform domain: %\small
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{-2ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which also includes language complement (\AF ∁).
\rnc\source{Calculus}
The \AF{ν} and \AF{δ} lemmas for codomain operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
(In contrast, the \AF{ν} and \AF{δ} lemmas do not easily generalize for domain operations.)
Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation with broader coverage code.

@@

\sectionl{Predicate Algebra}

@@

There is still more algebraic structure to be found and exploited.
When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Languages, Predicates, and Types}.
In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Languages, Predicates, and Types}.
The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
The proofs of these properties, while reasonably straightforward, are beyond the scope of this paper but are available in the paper's Agda source code.\out{\notefoot{There will be short and long forms of this paper. Perhaps the long form will contain narrated parts of the monoid semiring proof.}}

Not only do types form a closed semiring, the \emph{only} properties needed from types in the monoid semiring come from the laws of closed semirings.
One can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
\out{We will explore this generality in \secref{Beyond Predicates}.}
As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.

\subsectionl{Transporting Algebraic Properties}

\rnc\source{Transport}

Given that type-level languages have useful and interesting algebraic properties and that the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are tightly connected to languages, do the algebraic properties transfer to those implementations?
This question immediately raises a technical obstacle.
Algebraic abstractions and their properties are simply typed.
For instance, each instance of the \ARe{Monoid} abstraction involves a single type \AB τ with a binary operation {\AB{\_∙\_} \AK : \AB τ \AK → \AB τ \AK → \AB τ} and identity value {\AB ε \AK : \AB τ}.
This recipe accommodates the language operations in \figref{Lang-ops} and their generalizations in \figref{predicates} but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.

Fortunately, there is a simple way to encapsulate indexed types into simple (non-indexed) types, namely existential quantification.
For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
For a type family {\AB f \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB f} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB f \AB a}.
We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
For instance, consider language union from \figref{symbolic-api}:
\begin{code}
% Copied from latex/Symbolic.tex
\>[2]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∪\AgdaUnderscore{}}}%
\>[7]\AgdaSymbol{:}\AgdaSpace{}%
\AgdaDatatype{Lang}%
\>[15]\AgdaGeneralizable{P}%
\>[18]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaGeneralizable{Q}%
\>[28]\AgdaSymbol{→}\AgdaSpace{}%
\AgdaDatatype{Lang}\AgdaSpace{}%
\AgdaSymbol{(}\AgdaGeneralizable{P}%
\>[39]\AgdaOperator{\AgdaFunction{◬.∪}}%
\>[44]\AgdaGeneralizable{Q}\AgdaSymbol{)}\<%
\end{code}
Referring to the type and operations of \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'', define \agda{wrapped-Lang}
Then wrap union as follows:
\agda{wrapped-union}
Note that the type-level language components of the arguments and result are left to be inferred by the compiler.
\rnc\source{Existential}
Indexed operations of all arities for a type family \AB{f} can be systematically wrapped in this style, as shown in
\figrefdef{inj}{Existential wrappers}{\agda{inj}}, this time with the inferable components filled in for clarity.
Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.

In addition to wrapping types and operations, we also need to transport \emph{properties}.
A crucial question is the choice of equivalence relation for the new properties.
Choosing wrapped equivalence to be equivalence of \emph{first} projections (type \emph{indices}) results in all algebraic properties of those indices easily transporting to the existentially wrapped versions.
This choice may at first seem like cheating, since the entire operational aspect of the representation is ignored.
The correctness of that aspect, however, is guaranteed by its dependent typing, which anchors its meaning (no matter how cleverly implemented) to the non-operational type index.
For language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\agda{prop}}.
These definitions ease specification of algebraic instances for existentially wrapped types from corresponding instances for their indices.
% As an example, \figrefdef{wrap-semiring}{Lifting semiring instances}{\agda{Semiring}} 
As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift the \AF{\_∪\_}/\AF{\_⋆\_} language semiring of \secref{Specifying Languages} to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as follows (where \AF{\_≟\_} is decidable equality on list elements):\notefoot{Lift the \AF{\_∪\_}/\AF{\_∩\_} predicate semiring as well.}
\rnc\source{Transport}
\agda{examples}
